### 2023-04-24 chapter2/chapter2.tex
Moved rate and root figures by corresponding paragraphs in results
Added more points about scores analysis to conclusion outline
Wrote OU analysis introduction
Converted OU results to paragraphs and updated references

% RESULTS
% V. BM-OU analyses
%    A. While the the BM model can the rates of change of features after accounting the the phylogenetic relationships between sequences, they do not directly test for conservation of features
%         1. In fact, under the BM model, trait variation is unconstrained and will increase without bound over time
%         2. Instead, demonstrating evidence of conservation requires a comparison to a model where feature evolution is constrained.
%         3. A common choice is the Ornstein-Uhlenbeck (OU) model
%             i. Some explanation about how it works

@@

% III. Disorder score analyses
%    A. Comparison of features and disorder scores acts as a kind of ``mutational scanning'' experiment [citation to Eddy?]
%        1. Contrasts of features and disorder scores shows the disorder predictor is sensitive to most features used in this analyses
%        2. However, highest correlations to expected features like fraction_disorder and hydropathy
%        3. Interestingly, negatively correlated with wf_complexity. Many disordered regions have low levels of sequence complexity. However this is a statistical rather than a biophysical criterion, and some low complexity regions are structured, like collagen. This indicates that these models can confuse correlation and causation during training, and their representations should be interpreted carefully.
%        1. Furthermore, resolution of analyses limited because features calculated at level of regions rather than a fixed window around each residue.
%    B. GO annotation enrichment analysis suggests disorder scores undergo rapid evolution in proteins involved in DNA repair and found in extracellular structure
%        1. Not corrected for multiple testing nor is the false discovery rate controlled, so should be considered as hypotheses for further investigation
%        2. However, this result is at least consistent with other known functions of disorder
%            i. Interactions with extracellular components and signaling, both highly enriched in disorder?
%            ii. These regions could contain MoRFs, so they are already on the boundary between folded and disorder
%            iii. As evolution optimizes these interactions, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles
%            iv. The variable scoring by the disorder predictor may reflect this kind of ``phrase transition''

@@

\subsection{Brownian motion and Ornstein-Uhlenbeck analyses}
Brownian motion (BM) model parameters were calculated with two methods. The first used Felsenstein's contrasts algorithm to efficiently calculate roots and contrasts for the disorder scores and features of each region~\cite{Felsenstein1973, Felsenstein1981}. Rates were calculated as the mean of the squares of the contrasts. Though these values are unbiased, they are not maximum likelihood estimates and are inappropriate for use with the Akaike information criterion (AIC)~\cite{Akaike1974}. Thus, they were used for analyses involving only the BM model. For comparison with the Ornstein-Uhlenbeck (OU) model, the BM parameters were calculated by maximizing the likelihood. The OU model parameters were also calculated via maximum likelihood estimation. To avoid issues with model identifiability, the root was treated as a random variable~\cite{Ho2014}. Thus, the covariance matrix, $V$, was parameterized as $V_{ij} = e^{\alpha d_{ij}}$ where $d_{ij}$ is the tree distance between tips $i$ and $j$, and $\alpha$ is the selection strength~\cite{An2008}.

The AICs were calculated for the models of each feature using their maximized likelihoods and two and three parameters for the BM and OU models, respectively. The pairwise differences in the AICs yielded a vector with 82 components, each representing the relative goodness of fit of the OU model over the BM model after accounting for their difference in complexity. The vectors were clustered using the correlation distance metric and the UPGMA algorithm. Clusters were manually chosen for subsequent GO analyses. To enrich these clusters for regions with a high likelihood of feature conservation despite high levels of sequence divergence, this analysis was restricted to regions whose amino acid and indel substitution rates exceeded 1 and 0.1, respectively (Fig.~\ref{sfig:hierarchy_histogram}).

>>>>

\subsection{A model of constrained evolution reveals conserved features}
Though the BM process permits the inference of the rates of feature evolution after accounting for the phylogenetic relationships between species, it does not directly test for their conservation. In fact, under the BM model, trait variation is unconstrained and will increase without bound over time. Instead, evidence of conservation requires comparison to a model where trait variation is constrained. A common choice for modeling the effect of selection on the evolution of a continuous trait is the Ornstein-Uhlenbeck (OU) model. The OU model is similar to the BM model where a trait accumulates variation through small, undirected steps. However, it differs in the trait is also attracted towards an optimal value where the attraction is proportional to a parameter $\alpha$ and the trait's distance from this value. Under an additional assumption of stationarity that ensures parameter identifiability and estimate consistency, the OU model is specified with three total parameters: the optimal value, the fluctuation magnitude, and the selection strength~\cite{Ho2013, Ho2014}. While the first two parameters are analogous to the root and rate parameters in the BM model, respectively, the selection strength has no equivalent.

We therefore fit OU models to the features calculated from the segments in each region to detect evidence of their conservation. A strength of probabilistic models of trait evolution is they permit the use of standard statistical methods to compare the support for different hypotheses. Thus, the goodness of fit of the OU model relative to the BM model is a measure of the evidence for selective constraint. However, because the OU model is specified with an additional parameter and has a greater capacity to fit the data, we compared their fits using the difference in the Akaike information criterion (AIC), which accounts for the model complexity, rather than the likelihood ratio. We chose not to treat these differences in a strict statistical sense but instead as qualitative ``evolutionary signatures'' as in Zarin \textit{et al.}~\cite{Zarin2019}. We therefore clustered the signatures to identify patterns of feature conservation across the proteome. To ensure the clusters were enriched in regions with a high likelihood of feature conservation despite low levels of sequence identity, we restricted this analysis to regions with a minimum amount of divergence as measured by substitution models (Fig.~\ref{sfig:hierarchy_histogram}).

@@

% III. Disorder score analyses
%    A. Comparison of features and disorder scores acts as a kind of ``mutational scanning'' experiment [citation to Eddy?]
%        1. Contrasts of features and disorder scores shows the disorder predictor is sensitive to most features used in this analyses
%        2. However, highest correlations to expected features like fraction_disorder and hydropathy
%        3. Interestingly, negatively correlated with wf_complexity. Many disordered regions have low levels of sequence complexity. However this is a statistical rather than a biophysical criterion, and some low complexity regions are structured, like collagen. This indicates that these models can confuse correlation and causation during training, and their representations should be interpreted carefully.
%        4. Furthermore, resolution of analyses limited because features calculated at level of regions, whose lengths can exceed 1,000 residues, rather than a fixed window around each residue.
%        5. Disorder scores are weakly correlated with many motifs. Furthermore, unclear if predictor is responding directly to these motifs as a signal of disorder, or they appear in regions which evolving towards disorder. Interestingly docking (DOC) and ligand binding (LIG) sites tend to have negative correlations whereas the modification (MOD) sites have positive correlations. Docking and ligand binding sites are generally mediated by small hydrophobic patches, so the predictor could be responding to increased hydrophobicity caused by additional of hydrophobic residue ``completing'' the motif rather than the motif itself. Likewise, phosphorylation sites are highly enriched in IDRs and many are targeted to disorder-promoting residues like serine or threonine. Thus, it's unclear if the predictor is responding directly to these features or features which are correlated with them. More in depths of the representations of these disorder predictors would shed led on these questions.
%    B. GO annotation enrichment analysis suggests disorder scores undergo rapid evolution in proteins involved in DNA repair and found in extracellular structure
%        1. Not corrected for multiple testing nor is the false discovery rate controlled, so should be considered as hypotheses for further investigation
%        2. However, this result is at least consistent with other known functions of disorder
%            i. Interactions with extracellular components and signaling, both highly enriched in disorder?
%            ii. These regions could contain MoRFs, so they are already on the boundary between folded and disorder
%            iii. As evolution optimizes these interactions, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles
%            iv. The variable scoring by the disorder predictor may reflect this kind of ``phrase transition''

@@

\subsection{Brownian motion and Ornstein-Uhlenbeck analyses}
Brownian motion (BM) model parameters were calculated with two methods. The first used Felsenstein's contrasts algorithm to efficiently calculate roots and contrasts for the disorder scores and features of each region~\cite{Felsenstein1973, Felsenstein1981}. Rates were calculated as the mean of the squares of the contrasts. Though these values are unbiased, they are not maximum likelihood estimates and are inappropriate for use with the Akaike information criterion (AIC)~\cite{Akaike1974}. Thus, they were used for analyses involving only the BM model. For comparison with the Ornstein-Uhlenbeck (OU) model, the BM parameters were calculated by maximizing the likelihood. The OU model parameters were also calculated via maximum likelihood estimation. To ensure parameter identfiability and estimate consistency, the root was treated as a random variable~\cite{Ho2013, Ho2014}. Thus, the covariance matrix, $V$, was parameterized as $V_{ij} = e^{\alpha d_{ij}}$ where $d_{ij}$ is the tree distance between tips $i$ and $j$, and $\alpha$ is the selection strength~\cite{Ho2013}.

The AICs were calculated for the models of each feature using their maximized likelihoods and two and three parameters for the BM and OU models, respectively. The pairwise differences in the AICs yielded a vector with 82 components, each representing the relative goodness of fit of the OU model over the BM model after accounting for their difference in complexity. The vectors were clustered using the correlation distance metric and the UPGMA algorithm. Clusters were manually chosen for subsequent GO analyses. To enrich these clusters for regions with a high likelihood of feature conservation despite low levels of sequence identity, this analysis was restricted to regions whose amino acid and indel substitution rates exceeded 1 and 0.1, respectively (Fig.~\ref{sfig:hierarchy_histogram}).
